---
title: "The Languages of the World"
subtitle: "INFO 526 - Final Project"
author: 
  - name: "Harrison Horan"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Visualizing the TidyTuesday The Languages of the World dataset, curated from Glottolog 5.2.1, an open-access database in linguistics."
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
---

## Setup

```{r}
#| label: load-packages

pacman::p_load(tidyverse, 
               here, 
               countdown, 
               tidytuesdayR, 
               lubridate, 
               janitor, 
               showtext, 
               scico, 
               DT, 
               readr, 
               ggridges, 
               sf,
               ggmap, 
               leaflet, 
               mapproj, 
               geofacet, 
               geosphere,
               glue,
               splitstackshape,
               countrycode,
               rnaturalearth,
               patchwork,
               scales
               )

# theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))
options(width = 65)

# figure defaults
knitr::opts_chunk$set(
  fig.width = 9,
  fig.asp = 0.618,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)
```

# Question 1 - Global Language Patterns

What geographic patterns exist for endangered languages around the world?

## Part 1 - Endangered vs. Isolate Languages

What relationship exists between a language's isolate status and its endangered status?

##### What is a language isolate?

Per [Wikipedia](https://en.wikipedia.org/wiki/Language_isolate), a **language isolate**, sometimes called an **isolated language**, is a language that has no demonstrable genealogical relationship with any other language. That is, an isolate is a language family that only contains one member.

##### How are we defining 'endangered' languages?

The endangered statuses of languages, based on a six-point scale called the [Agglomerated Endangerment Scale (AES)](https://glottolog.org/langdoc/status), comes from [GlottoScope](https://glottolog.org/langdoc/status), which is a collection of global linguists dedicated to tracking endangered languages. The endangered statuses are as follows:

|   **AES status**   | AES Status Code | **\# of languages** | **% of languages** |
|:----------------:|:----------------:|:-----------------:|:----------------:|
| **Not Endangered** |        1        |        2770         |       35.34%       |
|   **Threatened**   |        2        |        1565         |       19.96%       |
|    **Shifting**    |        3        |        1799         |       22.95%       |
|    **Moribund**    |        4        |         421         |       5.37%        |
| **Nearly Extinct** |        5        |         300         |       3.83%        |
|    **Extinct**     |        6        |         984         |       12.55%       |
|     **total:**     |                 |        7839         |                    |

### Data Wrangling

#### Load Primary Data

```{r}
#| label: load-primary-data
#| message: False

# Download raw data and filter to endangered status
endangered_status <- 
  readr::read_csv("https://raw.githubusercontent.com/glottolog/glottolog-cldf/refs/heads/master/cldf/values.csv") |> 
  dplyr::filter(Parameter_ID == "aes") |> 
  dplyr::select(Language_ID, Value, Code_ID) |> 
  dplyr::rename(id = Language_ID,
                status_code = Value,
                status_label = Code_ID) |> 
  dplyr::mutate(status_label = stringr::str_replace(stringr::str_remove(status_label, "^aes-"), "_", " "))

# Download language and family data
fam_lgs <- 
  readr::read_csv("https://raw.githubusercontent.com/glottolog/glottolog-cldf/refs/heads/master/cldf/languages.csv")

# Filter and clean language family data
families <- 
  fam_lgs |> 
  dplyr::filter(Level == "family") |> 
  dplyr::select(ID, Name) |> 
  dplyr::rename(Family = Name) |> 
  dplyr::rename_with(stringr::str_to_lower, dplyr::everything())

# Filter and clean language data
languages <- 
  fam_lgs |> 
  dplyr::filter(Level == "language") |> 
  dplyr::select(ID, Name, Macroarea, Latitude, Longitude, ISO639P3code, Countries, Is_Isolate, Family_ID) |> 
  dplyr::rename_with(stringr::str_to_lower, dplyr::everything())
```

#### Prepare Data

```{r}
#| label: q1p1-wrangling
#| warning: False

# Create master languages data frame
master_languages <- left_join(
  languages,
  endangered_status,
  by = "id"
) |>
  left_join(
    families,
    by     = c("family_id" = "id"),
    suffix = c("", "_name")
  ) |>
  filter(
    if_all(c(
      latitude, longitude, countries, status_code, status_label
      ), ~ !is.na(.x))
    ) |>
  mutate(
    status_label = str_to_title(status_label),
    status_label = factor(status_label, levels = c(
    "Not Endangered", "Threatened", "Shifting", 
    "Moribund", "Nearly Extinct", "Extinct")),
    not_endangered = if_else(
      status_label == "Not Endangered", "Not Endangered", "Threatened or Worse")
    )
```

### Visualization

```{r}
#| label: q1p1-plotting


ggplot(master_languages, aes(
  x = status_label, 
  fill = is_isolate
  )) + 
  geom_bar() + 
  labs(
    title = "Count of Isolate Languages by Endangered Status",
    subtitle = "",
    x = "Endangered Status Designation",
    y = "Count of Languages",
    fill = "Is Isolate",
    caption = "Source: "
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_text(angle = 20)
  )
```

### Analysis & Discussion

## Part 2 - Largest Geographic Reach

Which language families span the largest geographic range?

Answering this question is only possible with a functional definition of 'largest geographic range'. I considered two different options:

1.  Haversine Distance, or essentially a point A to point B line between the two farthest apart speakers
2.  Bounding Box Area, which takes the min and max of latitude and longitude of speakers and creates a rectangle/area

### Data Wrangling

#### Calculate Custom Global Reach Metrics

```{r}
#| label: calculate-custom-metrics
#| message: False

# --- Option 1: Max Pairwise Haversine Distance ---
global_reach_haversine <- function(lons, lats) {
  coords <- cbind(lons, lats)
  if (nrow(coords) < 2) return(NA_real_)
  
  # distm returns a matrix of pairwise distances in meters
  dist_matrix <- geosphere::distm(coords, fun = distHaversine)
  max(dist_matrix, na.rm = TRUE) / 1000  # return in km
}

# --- Option 2: Bounding Box Area ---
global_reach_bbox <- function(lons, lats) {
  lat_range <- diff(range(lats, na.rm = TRUE))
  lon_range <- diff(range(lons, na.rm = TRUE))
  lat_range * lon_range  # in decimal degrees²
}


family_reach <- master_languages |>
  mutate(family = replace_na(family, "Isolate")) |>
  group_by(family) |>
  filter(!family %in% c(
  "Unattested", "Bookkeeping", "Unclassifiable",
  "Speech Register", "Pidgin", "Mixed Language"
  # "Artificial Language", "Sign Language", "Isolate"
  )) |>
  summarise(
    reach_km         = global_reach_haversine(longitude, latitude),
    reach_bbox_deg2  = global_reach_bbox(longitude, latitude),
    min_lat          = min(latitude, na.rm = TRUE),
    max_lat          = max(latitude, na.rm = TRUE),
    min_lon          = min(longitude, na.rm = TRUE),
    max_lon          = max(longitude, na.rm = TRUE),
    bb_asp           = diff(range(latitude, na.rm = TRUE)) / diff(range(longitude, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(reach_bbox_deg2))

top_10_reach <- family_reach |>
  slice_head(n = 10)
```

### Visualization

```{r}
#| fig-width: 10
#| fig-height: 16

# Create world map
world_map <- 
  ggplot() +
  geom_polygon(
    data = map_data("world"), 
    aes(x = long, y = lat, group = group),
    fill = "gray90", color = "white", linewidth = 0.1) + 
  labs(x = NULL, y = NULL) + 
  coord_map(
    projection = "mercator",
    xlim = c(-180, 180),
    ylim = c(-50, 70)
  ) +
  theme_minimal() + 
  theme(
    panel.grid   = element_blank(),
    panel.border = element_blank(),
    axis.text    = element_blank(),
    axis.ticks   = element_blank()
  )

world_map + 
  geom_rect(
    data = top_10_reach,
    aes(xmin  = min_lon, xmax = max_lon, 
        ymin  = min_lat, ymax = max_lat,
        color = family, 
        fill  = after_scale(alpha(color, 0.3))),
    linewidth = 0.8
  ) + 
  geom_hline(yintercept = 0, color = "darkgray", linewidth = 0.5, linetype = "dotted") + 
  geom_vline(xintercept = 0, color = "darkgray", linewidth = 0.5, linetype = "dotted") + 
  facet_wrap(~ family, ncol = 2) + 
  labs(
    title = "Global Reach of Top 10 Language Families",
    subtitle = "Total Area of Recorded Speakers",
    caption = "Source: 'The Languages of the World' TidyTuesday dataset, curated from Glottolog 5.2.1",
    color = "Family"
       ) +
  theme(
    legend.position = "none",
    plot.title      = element_text(hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5),
    panel.spacing   = unit(1, "cm")
  )
```

```{r}
ggplot(top_10_reach, aes(x = reorder(family, reach_km), y = reach_km)) +
  geom_col(fill = "#3d1a8e") +
  coord_flip() +
  labs(
    title = "Global Reach of Top 10 Language Families",
    subtitle = "By Longest Distance (km) Between Recorded Speakers",
    caption = "Source: 'The Languages of the World' TidyTuesday dataset, curated from Glottolog 5.2.1",
    x = NULL,
    y = "Haversine Line Distance (km)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
  )
```

### Analysis & Discussion

It's unconventional, and maybe crude, to use a rectangular area (calculated from extremes) to measure the global reach of a language family.

#### Challenges & Solutions

I think the biggest challenge with this visualization is semantic: language family designations don't really mean much to us. While most Americans might guess that Indo-European is 'what we speak', I don't think languages' families are common knowledge.

# Question 2 - Socioeconomic vs. Linguistic Change

1.  What relationship exists between shifting/endangered languages and:
    1.  Population change over the last 20 years
    2.  Change in GDP over the last 20 years

## Part 1 - Shifting Languages vs. Population Change

What relationship exists between shifting/endangered languages and the average population change of the countries in which it's spoken (last 20 years).

### Data Wrangling

#### Load External Source Data

```{r}
#| label: load-primary-data
#| message: False

# World Bank aggregate codes to exclude from gdp dataset
wb_aggregates <- c("AFE", "AFW", "ARB", "CEB", "CHI", "CSS", "EAP", "EAR", 
                   "EAS", "ECA", "ECS", "EMU", "EUU", "FCS", "HIC", "HPC", 
                   "IBD", "IBT", "IDA", "IDB", "IDX", "INX", "LAC", "LCN", 
                   "LDC", "LIC", "LMC", "LMY", "LTE", "MEA", "MIC", "MNA", 
                   "NAC", "OED", "OSS", "PRE", "PSS", "PST", "SAS", "SSA", 
                   "SSF", "SST", "TEA", "TEC", "TLA", "TMN", "TSA", "TSS", 
                   "UMC", "WLD", "XKX")

# Gross domestic product, reformatted
gdp <- 
  read_csv("data/GDP_by_country_time.csv", skip = 4) |>
  select(-`Indicator Name`, -`Indicator Code`) |>
  rename(country = `Country Name`, country_code = `Country Code`) |>
  filter(!country_code %in% wb_aggregates) |>
  select(where(~!all(is.na(.)))) |>  
  pivot_longer(
    cols = -c(country, country_code),
    names_to = "year",
    values_to = "gdp_growth_yoy"
  ) |>
  mutate(
    year = as.integer(year),
    gdp_growth_yoy = as.numeric(gdp_growth_yoy),
    country_code = countrycode(
      country_code, 
      origin = "iso3c", 
      destination = "iso2c",
      custom_match = c("XKX" = "XK")
    )
  ) |>
  filter(!is.na(gdp_growth_yoy)) |>
  filter(year >= 2004) |>
  arrange(country)

# Population change and density data by country
pop <- 
  read_csv("data/unpopulation_dataportal.csv") |>
  filter(
    Time >= 2004, 
    IndicatorName == "Rate of population change",
    Variant == "Median",
    Sex == "Both sexes"
  ) |>
  select("IndicatorName", "Location", "Iso2", "Iso3", "Time", "Value") |>
  rename(year = "Time", country = "Location", country_code = "Iso2") |>
  select(where(~!all(is.na(.)))) |> 
  pivot_wider(
    names_from = "IndicatorName",
    values_from = "Value"
  ) |>
  rename(pop_change_yoy = "Rate of population change") |>
  mutate(
    year = as.integer(year),
    pop_change_yoy = as.numeric(pop_change_yoy)
  )

# UN developing countries list
dev_countries <- 
  read_csv("data/un_developing_countries_data.csv") |>
  mutate(country_code = countrycode(
    country,
    origin = "country.name",                                
    destination = "iso2c",
    custom_match = c(
      "Democratic Rep. of the Congo"      = "CD",
      "Lao People's Dem. Republic"        = "LA",
      "Plurinational State of Bolivia"    = "BO",
      "Micronesia (Federated States of)"  = "FM",
      "São Tomé and Príncipe"             = "ST",
      "Timor-Leste"                       = "TL"
    )
  ))
```

#### Prepare Data

```{r}
#| label: q2p1-wrangling
#| warning: False

# Separating delimited countries & joining w/ custom datasets
lang_country_change <- master_languages |>
  filter(!is.na(countries)) |>
  select(-iso639p3code) |>
  separate_rows(countries, sep = ";") |>  # one row per country per language
  rename(country_code = "countries") |>
  mutate(
    status_code = as.integer(status_code)
  ) |>
  left_join(
    gdp,
    by = "country_code"
  ) |>
  left_join(
    pop |> select(-country, -Iso3),
    by = c("country_code", "year")
  )

# Separating delimited countries & joining w/ custom datasets
master_countries <- lang_country_change |>
  group_by(country, country_code, status_label, status_code) |>
  summarise(
    n_languages = n_distinct(id), 
    avg_annual_gdp_change_20_year = mean(gdp_growth_yoy, na.rm = TRUE),
    avg_annual_pop_change_20_year = mean(pop_change_yoy, na.rm = TRUE),
    .groups = "drop"
    ) |>
  group_by(country, country_code) |>
  mutate(
    total_languages   = sum(n_languages),
    status_proportion = n_languages / total_languages,
    avg_status_score  = mean(status_code)
  )

# Iteration of master countries with only shifting languages
master_countries_shifting <- master_countries |>
  filter(status_label == "Shifting")
```

### Visualization

```{r}
#| label: q2p1-plotting
#| warning: False

ggplot(master_countries, aes(x = avg_annual_gdp_change_20_year, 
                             y = status_label, 
                             color = status_label)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.3, size = 1) + 
  labs(
    title = "Distribution of Languages by Endangered Status and Language Country's 20 Year Avg. Annual Change in GDP"
  ) + 
  theme_minimal() + 
  theme(
    
  )

```

### Analysis & Discussion

## Part 2 - Shifting Languages vs. Shifting GDP

What relationship exists between shifting/endangered languages and average GDP change of the countries in which it's spoken (last 20 years).

### Prepare Data

```{r q2p2-data-wrangling}
#| message: False
#| warning: False

country_economics <- master_countries |>
  distinct(country, country_code, 
           avg_annual_gdp_change_20_year, 
           avg_annual_pop_change_20_year
           )

world <- ne_countries(scale = "medium", returnclass = "sf")

map_data_status <- world |>
  left_join(master_countries_shifting, by = c("iso_a2" = "country_code"))

map_data_econ <- world |>
  left_join(country_economics, by = c("iso_a2" = "country_code"))

map_dev_countries <- world |>
  left_join(dev_countries, by = c("iso_a2" = "country_code")) |>
  filter(iso_a2 %in% dev_countries$country_code)
```

### Visualization

```{r q2p2-plotting}

p1 <- ggplot(map_data_status) +
  geom_sf(aes(fill = status_proportion), color = "white", linewidth = 0.1) +
  geom_sf(data = map_dev_countries, fill = NA, color = "white", linewidth = 0.3) +
  coord_sf(ylim = c(-50, 85), expand = FALSE) +
  scale_fill_viridis_c(option = "rocket", direction = -1, na.value = "grey90") +
  labs(title = "Shifting Languages", fill = NULL) +
  theme_minimal()

p2 <- ggplot(map_data_econ) +
  geom_sf(aes(fill = avg_annual_gdp_change_20_year), color = "white", linewidth = 0.1) +
  coord_sf(ylim = c(-50, 85), expand = FALSE) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey90") +
  labs(title = "Shifting Economies", fill = NULL) +
  theme_minimal()

p3 <- ggplot(map_data_econ) +
  geom_sf(aes(fill = avg_annual_pop_change_20_year), color = "white", linewidth = 0.1) +
  coord_sf(ylim = c(-50, 85), expand = FALSE) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey90") +
  labs(title = "Shifting Populations", fill = NULL) +
  theme_minimal()

p4 <- ggplot(map_data_status) +
  geom_sf(aes(fill = status_proportion), color = "white", linewidth = 0.1) +
  geom_sf(data = map_dev_countries, fill = NA, color = "green", linewidth = 0.5) +
  coord_sf(ylim = c(-50, 85), expand = FALSE) +
  scale_fill_viridis_c(option = "rocket", direction = -1, na.value = "grey90") +
  labs(
    title = "Shifting Languages", 
    fill = "Proportion of Languages with\nan Endangered Status of 'Shifting'"
    ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
   # legend.position = "inside"
  )

p4
```

### Analysis & Discussion
