---
title: "INFO 526 Final Project Proposal"
subtitle: "The Languages of the World"
author: 
  - name: "Harrison Horan"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Visualizing the TidyTuesday The Languages of the World dataset, curated from Glottolog 5.2.1, an open-access database in linguistics."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(here)
library(countdown)

# theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))
options(width = 65)

# figure defaults
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

```

## Dataset

```{r}
#| label: load-dataset
#| message: false
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

The two questions you want to answer.

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
