---
title: "INFO 526 Final Project Proposal"
subtitle: "The Languages of the World"
author: 
  - name: "Harrison Horan"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Visualizing the TidyTuesday The Languages of the World dataset, curated from Glottolog 5.2.1, an open-access database in linguistics."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r setup}
#| label: load-pkgs
#| message: false

if (!require("pacman"))
  install.packages("pacman")

pacman::p_load(tidyverse, here, countdown, tidytuesdayR)

# theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))
options(width = 65)

# figure defaults
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

```

## Dataset - The Languages of the World

Source: <https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-12-23/readme.md>

```{r}
#| label: load-dataset
#| message: false

tuesdata <- tidytuesdayR::tt_load('2025-12-23')

# Primary datasets
endangered_status <- tuesdata$endangered_status
families <- tuesdata$families
languages <- tuesdata$languages

# Base, pre-cleaned data set underlying families and languages sets
fam_lgs <- 
  readr::read_csv("https://raw.githubusercontent.com/glottolog/glottolog-cldf/refs/heads/master/cldf/languages.csv")
```

There are three primary datasets from TidyTuesday, each of which are derived from [Glottolog](https://glottolog.org/), which is an open-access linguistics database.

The focus of the data are the languages spoken across the world, where they are spoken (both in terms of macroarea and lat/lon coordinates), and which families they belong to. Beyond that, and maybe the most pointed data, are the endangered statuses of these languages, based on a six-point scale called the [Agglomerated Endangerment Scale (AES)](https://glottolog.org/langdoc/status), which is as follows:

|   **AES status**   | AES Status Code | **\# of languages** | **% of languages** |
|:--------------------:|:---------------:|:---------------:|:---------------:|
| **not endangered** |        1        |        2770         |       35.34%       |
|   **threatened**   |        2        |        1565         |       19.96%       |
|    **shifting**    |        3        |        1799         |       22.95%       |
|    **moribund**    |        4        |         421         |       5.37%        |
| **nearly extinct** |        5        |         300         |       3.83%        |
|    **extinct**     |        6        |         984         |       12.55%       |
|     **total:**     |                 |        7839         |                    |

Note: The 'families' and 'languages' datasets are derived from 'fam_lgs', using the following script provided by TidyTuesday:

``` {.r fold="True"}
# Imports
library(tidyverse)

# Download raw data and filter to endangered status
endangered_status <- 
  readr::read_csv("https://raw.githubusercontent.com/glottolog/glottolog-cldf/refs/heads/master/cldf/values.csv") |> 
  dplyr::filter(Parameter_ID == "aes") |> 
  dplyr::select(Language_ID, Value, Code_ID) |> 
  dplyr::rename(id = Language_ID,
                status_code = Value,
                status_label = Code_ID) |> 
  dplyr::mutate(status_label = stringr::str_replace(stringr::str_remove(status_label, "^aes-"), "_", " "))

# Download language and family data
fam_lgs <- 
  readr::read_csv("https://raw.githubusercontent.com/glottolog/glottolog-cldf/refs/heads/master/cldf/languages.csv")

# Filter and clean language family data
families <- 
  fam_lgs |> 
  dplyr::filter(Level == "family") |> 
  dplyr::select(ID, Name) |> 
  dplyr::rename(Family = Name) |> 
  dplyr::rename_with(stringr::str_to_lower, dplyr::everything())

# Filter and clean language data
languages <- 
  fam_lgs |> 
  dplyr::filter(Level == "language") |> 
  dplyr::select(ID, Name, Macroarea, Latitude, Longitude, ISO639P3code, Countries, Is_Isolate, Family_ID) |> 
  dplyr::rename_with(stringr::str_to_lower, dplyr::everything())
```

## Questions

1.  What geographic patterns exist for endangered languages around the world?
2.  Where is language shifting the most in the world?

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

### Question 1 - Global Patterns

### Question 2 - Shifting Regions
